Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-06-13 paper/document.tex
Turd polishing.

In this paper, we propose a method that analyses the changes in type usage and method invocations over time. Using commits, we are able to keep track of \emph{who} is changing what types. We then present this data using simple visualizations. We claim that type usage and method invocation offers more insight than line-based, file-based, and commit-based methods of visualization and analysis. For simplicity, we analyze only Java projects, as it allowed us to quickly develop and test our software. However, since the approach described in this paper can be generally applied to any statically-typed language, it can be extended it to other languages. Our goal is to describe a methodology for obtaining richer, more meaningful data from source code repositories which can be used in more insightful analyses.

@@

Further analysis of the Java language has been done using ASTs. L\"{a}mmel \textit{et al.}~\cite{Lammel:2011:LAA:1982185.1982471} used ASTs to examine the usage of APIs in Java projects. They found what APIs are popular and if they were used in a framework-like manner. Dyer \textit{et al.}~\cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. The authors found the most popular features and the adoption rate of new features over time. They did not check to see how much a developer used a feature but instead checked to see if they used it at all. In a more general analysis of programming languages, Meyerovich and Rabkin \cite{Meyerovich:2013:EAP:2509136.2509515} surveyed developers and examined repositories to learn about language adoption and usage. Developers feel that certain language features are more important than others.

@@

The basic tools easily available to any software developer are the GitHub graphs (see Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. GitHub tracks contributions based only on the number of commits and number of lines changed. This has the advantage of not relying on the type of code which was committed. It does not matter what language is used or if errors exist in the commit, GitHub will simply output the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear idea of is being changed, only that changes have occurred. This data may be lacking, or even misleading which contributes to the need for better tools.

@@

To determine what has changed between two subsequent commits, we computed their \emph{AST difference}. Given two ASTs, an AST difference (or simply \emph{AST diff}) states which nodes were must be added and which nodes must be removed to transform one tree to the next (step two in Figure~\ref{fig:context}). This allows us to precisely track changes two consecutive revisions of the same file. For AST diffing, we used GumTree~\cite{falleri:hal-01054552}, an algorithm made to compute the difference between two ASTs generated by Spoon. For new files and deleted files, it is unnecessary to calculate an AST difference; we simply run Spoon on the files and treated everything as an addition or deletion respectively.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit, we collected the following information, as can be seen in Figure~\ref{fig:data-extracted}: the author's name, date of commit, commit ID (SHA), commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited Java files we also compared the file at the time of the commit with its parent commit using AST differencing as described above. We achieved this by using calling \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent, such as the initial commit. We were able to handle this by calling \texttt{git show} and saving the output to a temporary file to be used with the AST differencing tool. When calling \texttt{git show} instead of \texttt{git difftool} we treat everything as an addition, as in commit A in Figure~\ref{fig:commits}. Since we compared each commit to its parent, as reported by Git, we were able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, like commit D in Figure~\ref{fig:commits}, it indicates a merge commit in Git. Merge commits contain the changes made in each parent commit combined. Therefore, we ignored merge commits since all changes in the merge should already have been accounted for in previous commits. Thus, any branch that has been merged into the primary branch would have been analyzed and included by TypeV.

@@

TypeV provides a variety of views to help organize the data and provide different levels of detail. The overview displays several rows of stacked bar graphs which show changes to a type or method over a period of time in a project (Figure~\ref{fig:main-view}). The purpose of displaying stacked bar charts is that they are a familiar visualization, thus are somewhat easy to read, and can succinctly display the relative amounts of additions and deletions per each time period. The light blue \additionsbox{10pt} and dark red \deletionsbox{10pt} portions account for the number of additions and deletions respectively, in a similar fashion to GitHub's code frequency graph (Figure~\ref{fig:GH_code_freq}). These colours were chosen as they are high contrast, and are distinct to those red-green colour blindness.

@@

If we tried to do the same analysis with the GitHub tools it would be significantly more difficult to achieve the conclusion achieved using TypeV (Figure~\ref{fig:parser2}). First if we only look at line changes this change to \texttt{ParserATNSimulator} is overwhelmed by other changes and hardly shows up on the GitHub tools. If we did notice an increase in activity we would not know what types were being worked on. Finally, we could not quickly look at the commit messages for that specific time and type if we used the GitHub tools.

@@

Bug triaging for large open source projects is a very time consuming task~\cite{badashian2015}, with hundreds  of bugs being reported daily. Each bug must be processed and assigned to a developer that has the experience required to fix the bug. In order to be effective, the triager has to have knowledge of what each developer has worked on as well as the parts of the code may be affected by the bugs. With TypeV, the triager is able to directly see what types a developer has worked with over time. This is important information for triaging since the type causing the bug, if known, will have a list of associated developers who have recently used that type and can be assigned the bug. The commit messages for the affected types are also visible, providing greater context about how the type has been used. The assigned developer can also use type information to locate other pieces of code affected by a bug and know which developers have been using the affected types.

>>>>

In this paper, we propose a method that analyses the changes in type usage and method invocations over time. Using commits, we are able to keep track of \emph{who} is changing what types. We then present this data using simple visualizations. We claim that type usage and method invocation offers more insight than line-based, file-based, and commit-based methods of visualization and analysis. For simplicity, we analyze only Java projects, as it allowed us to quickly develop and test our software. However, since the approach described in this paper can be generally applied to any statically-typed language, it can be extended to other languages. Our goal is to describe a methodology for obtaining richer, more meaningful data from source code repositories which can be used in more insightful analyses.

@@

Further analysis of the Java language has been done using ASTs. L\"{a}mmel \textit{et al.}~\cite{Lammel:2011:LAA:1982185.1982471} used ASTs to examine the usage of APIs in Java projects. They found which APIs are popular and if they were used in a framework-like manner. Dyer \textit{et al.}~\cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. The authors found the most popular features and the adoption rate of new features over time. They did not check to see how much a developer used a feature but instead checked to see if they used it at all. In a more general analysis of programming languages, Meyerovich and Rabkin \cite{Meyerovich:2013:EAP:2509136.2509515} surveyed developers and examined repositories to learn about language adoption and usage. Developers feel that certain language features are more important than others.

@@

The basic tools easily available to any software developer are the GitHub graphs (see Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. GitHub tracks contributions based only on the number of commits and number of lines changed. This has the advantage of not relying on the type of code which was committed. It does not matter what language is used or if errors exist in the commit, GitHub will simply output the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear idea of what is being changed, only that changes have occurred. This data may be lacking, or even misleading which contributes to the need for better tools.

@@

To determine what has changed between two subsequent commits, we computed their \emph{AST difference}. Given two ASTs, an AST difference (or simply \emph{AST diff}) states which nodes must be added and which nodes must be removed to transform one tree to the next (step two in Figure~\ref{fig:context}). This allows us to precisely track changes between two consecutive revisions of the same file. For AST diffing, we used GumTree~\cite{falleri:hal-01054552}, an algorithm made to compute the difference between two ASTs generated by Spoon. For new files and deleted files, it is unnecessary to calculate an AST difference; we simply run Spoon on the files and treated everything as an addition or deletion respectively.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit, we collected the following information, as can be seen in Figure~\ref{fig:data-extracted}: the author's name, date of commit, commit ID (SHA), commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited Java files we also compared the file at the time of the commit with its parent commit using AST differencing as described above. We achieved this by invoking \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent, such as the initial commit. We were able to handle this by calling \texttt{git show} and saving the output to a temporary file to be used with the AST differencing tool. When calling \texttt{git show} instead of \texttt{git difftool} we treat everything as an addition, as in commit A in Figure~\ref{fig:commits}. Since we compare each commit to its parent as reported by Git, we are able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, like commit D in Figure~\ref{fig:commits}, it indicates a merge commit in Git. Merge commits contain the changes made in each parent commit combined. Therefore, we ignored merge commits since all changes in the merge should already have been accounted for in previous commits. Thus, any branch that has been merged into the primary branch would have been analyzed and included by TypeV.

@@

TypeV provides a variety of views to help organize the data and provide different levels of detail. The overview displays several rows of stacked bar graphs which show changes to a type or method over a period of time in a project (Figure~\ref{fig:main-view}). The purpose of displaying stacked bar charts is that they are a familiar visualization, thus are somewhat easy to read, and can succinctly display the relative amounts of additions and deletions per each time period. The light blue \additionsbox{10pt} and dark red \deletionsbox{10pt} portions account for the number of additions and deletions respectively, in a similar fashion to GitHub's code frequency graph (Figure~\ref{fig:GH_code_freq}). These colours were chosen as they are high contrast, and are distinct to those experiencing red-green colour blindness.

@@

If we tried to do the same analysis with the GitHub tools it would be significantly more difficult to reach this conclusion (Figure~\ref{fig:parser2}). First, if we only look at line changes this change to \texttt{ParserATNSimulator} is overwhelmed by other changes and hardly shows up on the GitHub tools. If we did notice an increase in activity we would not know what types were being worked on. Finally, we could not quickly look at the commit messages for that specific time and type if we used the GitHub tools.

@@

Bug triaging for large open source projects is a very time consuming task~\cite{badashian2015}, with hundreds  of bugs being reported daily. Each bug must be processed and assigned to a developer that has the experience required to fix the bug. In order to be effective, the triager has to have knowledge of what each developer has worked on as well as the parts of the code that may be affected by the bugs. With TypeV, the triager is able to directly see what types a developer has worked with over time. This is important information for triaging since the type causing the bug, if known, will have a list of associated developers who have recently used that type and can be assigned the bug. The commit messages for the affected types are also visible, providing greater context about how the type has been used. The assigned developer can also use type information to locate other pieces of code affected by a bug and know which developers have been using the affected types.
